---
title: "CourseP2"
output: html_document

---

# Analysing the impact of severe weather events on health and economy

## Synopsis

This is a second course project for Reproducible Research course which is part of the Coursera’s Data Science Specialization.

Storms and other severe weather events are responsible for both public health and economic problems for communities and municipalities. Many severe weather events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database to find out what are the most damaging weather events. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.



## Data Processing

Loading the dataset(assuming it is placed in the same directory as this document).
```{r}
require(dplyr)
require(ggplot2)
require(gridExtra)
weatherDF <- read.csv("repdata_data_StormData.csv")
summary(weatherDF)
```

### PART 1 - Effects on public heatlh

Table with injuries and fatalities for each event, as well as the number of occurrences.

```{r}
healthTableSum <- cbind(aggregate(FATALITIES ~ EVTYPE, data=df, sum), aggregate(INJURIES ~ EVTYPE, data=df, sum)$INJURIES)
countTable <- df %>% count(EVTYPE)
healthTable <- cbind(healthTableSum,countTable[2])
colnames(healthTable)[2] <- "FATALITIES"
colnames(healthTable)[3] <- "INJURIES"
colnames(healthTable)[4] <- "OCURRENCES"


```

### PART 2 - Effects on the economy

Table with the sum of crop and property damage for each event.

```{r}
damageTable <-weatherDF[which(weatherDF$CROPDMGEXP != "" | weatherDF$PROPDMGEXP != ""),]
damageTable[which(damageTable$CROPDMGEXP =="B" | damageTable$CROPDMGEXP =="b" ),]$CROPDMG <- sapply(damageTable[which(damageTable$CROPDMGEXP =="B" | damageTable$CROPDMGEXP =="b"),]$CROPDMG,"*",1000000000)
damageTable[which(damageTable$CROPDMGEXP =="M" | damageTable$CROPDMGEXP =="m" ),]$CROPDMG <- sapply(damageTable[which(damageTable$CROPDMGEXP =="M" | damageTable$CROPDMGEXP =="m"),]$CROPDMG,"*",1000000)
damageTable[which(damageTable$CROPDMGEXP =="K" | damageTable$CROPDMGEXP =="k"),]$CROPDMG <- sapply(damageTable[which(damageTable$CROPDMGEXP =="K" | damageTable$CROPDMGEXP =="k"),]$CROPDMG,"*",1000)
damageTable <-weatherDF[which(weatherDF$PROPDMG > 0 & weatherDF$PROPDMGEXP != "" ),]
damageTable[which(damageTable$PROPDMGEXP =="B" | damageTable$PROPDMGEXP =="b" ),]$PROPDMG <- sapply(damageTable[which(damageTable$PROPDMGEXP =="B" | damageTable$PROPDMGEXP =="b"),]$PROPDMG,"*",1000000000)
damageTable[which(damageTable$PROPDMGEXP =="M" | damageTable$PROPDMGEXP =="m" ),]$PROPDMG <- sapply(damageTable[which(damageTable$PROPDMGEXP =="M" | damageTable$PROPDMGEXP =="m"),]$PROPDMG,"*",1000000)
damageTable[which(damageTable$PROPDMGEXP =="K" | damageTable$PROPDMGEXP =="k"),]$PROPDMG <- sapply(damageTable[which(damageTable$PROPDMGEXP =="K" | damageTable$PROPDMGEXP =="k"),]$PROPDMG,"*",1000)
damageTable <- cbind(aggregate(CROPDMG ~ EVTYPE, data=df, sum), aggregate(PROPDMG ~ EVTYPE, data=df, sum)$PROPDMG)
colnames(damageTable)[3] <- "PROPDMG"

```
## Results

### PART 1 - Effects on public health

First, let's analyze the 10 most fatal and injury making weather events.

```{r}
par(mfrow = c(3, 1), mar = c(7, 8, 2, 1), oma = c(0, 0, 2, 0),mgp = c(6, 1, 0))
with(head(healthTable[order(-healthTable$OCURRENCES),], 10),barplot(OCURRENCES,names.arg = EVTYPE,ylab= "Ocurrences",,col = "blue", las = 2,cex.names = 0.7))
with(head(healthTable[order(-healthTable$FATALITIES),], 10),barplot(FATALITIES,names.arg = EVTYPE,ylab= "Fatalities",col = "red", las = 2,cex.names = 0.7))
with(head(healthTable[order(-healthTable$INJURIES),], 10),barplot(INJURIES,names.arg = EVTYPE,xlab = "Events",ylab= "Injuries",col = "orange", las = 2,cex.names = 0.7))
mtext("10 most damaging events to public health regarding fatalities and injuries", side = 3, outer = TRUE)
```


From this graph, we can see that the tornadoes are by far the most health disruptive phenomenon, as it is the one with the most fatalities and injuries, while also being one of the most recurrent events. Excessive heat, floods and thunderstorms are the other most heath damaging events.

### PART 2 - Effects on the economy
```{r}
par(mfrow = c(3, 1), mar = c(7, 8, 2, 1), oma = c(0, 0, 2, 0),mgp = c(6, 1, 0))
with(head(healthTable[order(-healthTable$OCURRENCES),], 10),barplot(OCURRENCES,names.arg = EVTYPE,,ylab= "Ocurrences",xpd = FALSE,col = "blue",las = 2,cex.names = 0.7))
with(head(damageTable[order(-damageTable$CROPDMG),], 10),barplot(CROPDMG,names.arg = EVTYPE,,ylab= "Crop damage",xpd = FALSE,col = "red",las = 2,cex.names = 0.7))
with(head(damageTable[order(-damageTable$PROPDMG),], 10),barplot(PROPDMG,names.arg = EVTYPE,,ylab= "Property damage",xpd = FALSE,col = "orange",xlab = "Events",las = 2,cex.names = 0.7))
mtext("10 most damaging events to the economy regarding crop and property damage", side = 3, outer = TRUE)

```

From this graph, we can conclude that the most damaging events to the economy are tornados, hails, floods and thunderstorm winds.
